---
title: "Types of variables"
date: "StatPREP Class Lesson"
description: "Introduces the distinction between quantitative and categorical variables, the *range* of a quantitative variable and the *levels* of a categorical variable."
weight: 2
output:
  word_document:
    reference_docx: lesson-style.docx
  pdf_document: tint::tintPdf
---


```{r include = FALSE}
SDSdata::sds_setup()
library(LittleApp)
library(DT)
```

# Orientation

There are two major types of variables:

* **Quantitative**, where the *value* of the variable is a number.
* **Categorical**,  where the value of the variable is one of a set of labels. That is, the value tells which *category* or *group* the row falls into.

It's usually easy to tell what type a variable is by looking at the variable's column in the data frame. But it's also easy to tell from a point plot what are the types of the response and explanatory variables.

Quantitative variables cover an *interval* of the number line. That interval is described by two numbers: the minimum value and the maximum value of the variable.

Quantitative variables also have a *spread*. There are different ways to measure the spread. In this lesson, we'll use the *difference*  between the maximum and minimum value. Note that the *interval* is two numbers while the *spread* is a single number.

The *spread* is said to measure the *variability* in the variable. Other ways to quantify the spread, which we won't use in this lesson, are the *standard deviation* and the *variance*. 

For categorical variables, we don't use the concepts *interval* or *spread*. The possible or allowed values of a categorical variable are called the *levels* of that variable. For example, the levels of a categorical variable describing "commute type" might be: walk, bike, bus, drive, etc. The levels of a categorical variable like "language spoken" might be English, Spanish, Chinese, and so on.

The levels of many categorical variables are *unordered*. This means that the concept of *between* doesn't naturally apply. For instance, there is no  natural way that Spanish is between English and Chinese.

Some categorical variables are *ordinal*, which  is just to say that there is a natural order  to  the levels. An example is a variable recording "opinion," which might have levels Disagree, Neutral, Agree.

# Activity

```{r echo = FALSE}
point_plot_app <- get_app_info("point_plot")
```


Open the [Point Plot Little App](`r point_plot_app$url`). Set the data frame to `NHANES` and the sample size to be several hundred.

1. Pick a quantitative response variable and a categorical explanatory  variable. You might have to use trial and error to find such variables but once you do, it will be evident in the graph. Please don't use `ID` or `SurveyYr` -- stick to the variables that are about the individual being surveyed.

    **Tip**: When one or both of the response and explanatory variables is  categorical, it is often nice to *jitter* the point plot. Jittering doesn't change the values of the variable itself, it just changes how they are displayed in the point plot. Jittering moves each point a small random distance from the exact tick which marks the level of the variable along the axis.

    * Write down the names of the variables you selected. For the categorical variable, write down each of the levels.

    * Measure the *interval* spanned by the quantitative variable for each group defined by  the categorical variable. You can use the measuring stick built into the app. (That is,  click at a point in the graphics frame and drag the mouse to select a vertical interval.) Write down the interval for each categorical level.

    * Also, measure and record the *spread* of the quantitative variable for each categorical level. 

2. Using the selector for the explanatory variable, find five more categorical variables in the `NHANES` data. At least two of your five should be *ordinal* variables.

    * For each, write down the levels,  and whether the variable is unordered or ordinal.
    

----------------------

Version 0.2, `r Sys.Date()`, Daniel Kaplan, [Word version](`r paste0("/word-versions/",tools::file_path_sans_ext(current_input()), ".docx")`) 